---
title: "GI mapping pipeline: filter counts and calculate LFC"
author: "Phoebe Parrish"
date: "`r format(Sys.time(), '%m/%d/%y %I:%M %p')`"
output: html_document
params: 
  cell_line: "HeLa"
  base_dir: "/Volumes/Projects/paralog_pgRNA/pgPEN_library/GI_mapping"
---

## Rmd knit notes
https://stackoverflow.com/questions/32479130/passing-parameters-to-r-markdown
https://stackoverflow.com/questions/31463143/pass-parameters-from-command-line-into-r-markdown-document


## To Do
* add steps to filter out screens with SSMD < -0.5 or replicate correlation <0.7? 
* use only expressed pan-essential genes as + controls? 
* add "Rep" back into replicate labels
* add in early vs. late LFC? adjust as well? 
* clean up tibble earlier - get rid of extra unnecessary columns so it's more readable, drop old columns when necessary
* get broad_target_type back in?
* make plots that I want to do before & after adjusting functions? 
* make target mean expression density plots for single- and double-targeting pgRNAs
* flag if reps have SSMD > -0.5 & add in an option for user to re-run and remove them


## Setup
```{r, include = FALSE, eval = FALSE}

## possible params for future use: 
# params: 
#   counts_file: 
#     value: x
#   output_dir: 
#     value: x
# knit: (function(inputFile, encoding) {
#     rmarkdown::render(inputFile, 
#       encoding = encoding, 
#       output_dir = "`r params$output_dir`")
#   })
# knit: function(inputFile, encoding){
#   rmarkdown::render(inputFile, 
#     encoding = encoding, 
#     output_dir = "../results/pgRNA_counts_QC")}

```




### Load packages
```{r setup, results = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(tidylog)
library(RColorBrewer) # for heatmap colors
library(kableExtra) # for formatting kables
library(corrr)

```

### Save variables and functions

```{r}

## get functions and vars from shared R script
source(file.path(params$base_dir, "workflow", "scripts", "shared_functions_and_variables.R"))

contour_palette <- colorRampPalette(brewer.pal(n = 9, name ="Spectral"))(50)

```

```{r global_options}

## knitr options
# source(file.path(params$base_dir, "workflow", "scripts", "shared_Rmd_settings.R"))

```

```{r}

## file paths
in_dir <- file.path(params$base_dir, "results", "pgRNA_counts_QC")

annot_dir <- file.path(params$base_dir, "results", "pgRNA_annotations")

out_dir <- file.path(params$base_dir, "results", "calculate_LFC")

make_out_dir(out_dir)

```

### Save functions
```{r}

make_norm_ctrl_violin_plot <- function(df, y_var, y_lab){
  
  plot <- ggplot(df, aes(x = norm_ctrl_flag, y = get(y_var), fill = norm_ctrl_flag)) +
    geom_hline(yintercept = 0) +
    geom_violin() +
    geom_boxplot(outlier.shape = NA, coef = 0, width = 0.1) +
    labs(x = "pgRNA_category", y = y_lab) +
    plot_options +
    plot_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          aspect.ratio = wide_ar) 
    
  return(plot)
  
}

```

```{r}

get_rep_cor_dfs <- function(df, group_var, plot_var){
  
  d.reps <- df %>%
  ungroup() %>%
  dplyr::select(rep) %>%
  distinct()

  d.var_rep_cor <- df %>%
    ungroup() %>%
    dplyr::select(group_var, rep, plot_var) %>%
    pivot_wider(names_from = rep,
                values_from = plot_var) 

  d.cor <- d.var_rep_cor %>%
    dplyr::select(-group_var) %>%
    corrr::correlate() %>%
    shave() %>%
    stretch() %>%
    filter(!is.na(r)) %>%
    unite(c(x, y), col = "comparison", sep = "_vs_", remove = FALSE) %>%
    rename("sample1" = x, "sample2" = y)


  comparisons <- d.cor %>% pull(comparison)

  results <- lapply(comparisons, function(i){

    sample1 <- d.cor %>% filter(comparison == i) %>% pull(sample1)
    sample2 <- d.cor %>% filter(comparison == i) %>% pull(sample2)

    d.comparison_var <- d.var_rep_cor %>%
      dplyr::select(sample1, sample2) %>%
      rename("sample1" = sample1, "sample2" = sample2) %>%
      mutate("comparison" = i)
  })
  
  d.var_rep_cor_plot <- bind_rows(results)
  
  out_list <- list("plot_df" = d.var_rep_cor_plot, "cor_df" = d.cor)
  return(out_list)
  
}

```


```{r}

make_rep_cor_plot <- function(plot_df, cor_df, axis_label){

  plot <- ggplot(plot_df, aes(x = sample1, y = sample2)) + 
    geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray") +
    geom_point(size = 1) +
    geom_density2d(aes(color = ..level..), alpha = 0.7) +
    geom_text(data = cor_df, 
              mapping = aes(x = -Inf, y = Inf, label = paste0("r=", round(r, 3))), 
              hjust = -0.25, vjust = 1.75, size = 3.5) +
    scale_colour_gradientn(colors = rev(contour_palette)) +
    labs(x = paste("sample1", axis_label, sep = "_"),
         y = paste("sample2", axis_label, sep = "_"), 
         color = "density") +
    plot_options +
    plot_theme +
    theme(aspect.ratio = square_ar) +
    facet_wrap(~comparison)
  
  return(plot)
  
}


```


### Read in files

```{r, results = FALSE, message = FALSE, warning = FALSE}

## counts file, flagged with whether or not to keep each pgRNA
d.counts_flagged <- read_rds(file.path(in_dir, "tables", "rds", paste0("d.", params$cell_line, "_counts_cpm_flag_long")))

## annotations file
d.annot <- read_rds(file.path(annot_dir, "tables", "rds", paste0("d.", params$cell_line, "_annot")))

```


## Calculate LFC

### Reformat d.counts
```{r}

## calculate LFC
d.counts_flagged_timepoint <- d.counts_flagged %>%
  separate(sample, into = c("day", "rep"), sep = "_", remove = FALSE) %>%
  mutate(day = readr::parse_number(day),
         rep = str_replace(string = rep, pattern = "Rep", replacement = "")) %>%
  group_by(id) %>%
  mutate(timepoint = case_when(
    day == min(day) ~ "plasmid",
    day == max(day) ~ "late",
    TRUE ~ "early")) %>%
  ungroup()

```

### Calculate LFC
```{r}

## early timepoints
## this will have to be changed if there are multiple early TP reps
d.counts_flagged_timepoint_early <- d.counts_flagged_timepoint %>%
  filter(timepoint == "plasmid" | timepoint == "early") %>%
  dplyr::select(id, timepoint, log2_cpm) %>%
  pivot_wider(names_from = timepoint, 
              values_from = log2_cpm,
              names_glue = "{timepoint}_{.value}")
d.counts_flagged_timepoint_early

## calculate LFC for plasmid vs. late and early vs. late
d.counts_flagged_timepoint_lfc <- d.counts_flagged_timepoint %>%
  filter(timepoint == "late") %>% ## filter for only late timepoints
  left_join(d.counts_flagged_timepoint_early, by = "id") %>%
  mutate(lfc_plasmid_vs_late = log2_cpm - plasmid_log2_cpm,
         lfc_early_vs_late = log2_cpm - early_log2_cpm)
d.counts_flagged_timepoint_lfc

```

## Annotate & filter
```{r}

d.lfc_annot <- d.counts_flagged_timepoint_lfc %>%
  left_join(d.annot, by = c("id" = "pgRNA_id")) %>%
  rename("pgRNA_id" = id)

# d.lfc_annot

d.removed_pgRNAs <- d.lfc_annot %>%
  dplyr::select(pgRNA_id, paralog_pair, low_plasmid_cpm, zero_count, rm_pgRNA) %>%
  distinct(pgRNA_id, .keep_all = TRUE) %>%
  filter(rm_pgRNA == TRUE) 
save_tbl(d.removed_pgRNAs)

## filter out pgRNAs to remove
d.lfc_annot <- d.lfc_annot %>%
  filter(rm_pgRNA == FALSE)

```


## Before adjusting LFC
```{r}

# d.lfc_annot %>%
#   ggplot(aes(x = norm_ctrl_flag, y = lfc_plasmid_vs_late, fill = norm_ctrl_flag)) +
#   geom_hline(yintercept = 0) +
#   geom_violin() +
#   geom_boxplot(outlier.shape = NA, coef = 0, width = 0.1) +
#   labs(x = "pgRNA_category") +
#   plot_options +
#   plot_theme +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1),
#         legend.position = "none",
#         aspect.ratio = wide_ar) +
#   facet_wrap(~rep)

unadjusted_norm_ctrl_violin_plot <- make_norm_ctrl_violin_plot(d.lfc_annot, "lfc_plasmid_vs_late", "raw_LFC")

unadjusted_norm_ctrl_violin_plot +
    facet_wrap(~rep)

```

Note that negative control pgRNAs will appear to be positively selected. This is due to the fact that essential pgRNAs are dropping out. To account for this, we will adjust the LFC values to set the median of negative control pgRNAs to zero. 


### Calculate SSMD
```{r}

# d.lfc_annot

d.lfc_annot %>%
  filter(norm_ctrl_flag == "negative_control" | norm_ctrl_flag == "positive_control") %>%
  group_by(rep, norm_ctrl_flag) %>%
  summarize(mean = mean(lfc_plasmid_vs_late),
            sd = sd(lfc_plasmid_vs_late)) %>%
  pivot_wider(names_from = norm_ctrl_flag, 
              values_from = c(mean, sd),
              names_glue = "{norm_ctrl_flag}_{.value}") %>%
  mutate(ssmd = (positive_control_mean - negative_control_mean) /
           (sqrt(positive_control_sd^2 + negative_control_sd^2)))

d.lfc_annot %>%
  ungroup() %>%
  filter(norm_ctrl_flag == "negative_control" | norm_ctrl_flag == "positive_control") %>%
  mutate(keep_flag = case_when(
    target_type == "ctrl_ctrl" ~ TRUE,
    target_type == "gene_ctrl" & gene1_expressed_flag == TRUE ~ TRUE,
    target_type == "ctrl_gene" & gene2_expressed_flag == TRUE ~ TRUE,
    TRUE ~ FALSE)) %>% ## if these conditions are not met, set to F
  filter(keep_flag == TRUE) %>%
  group_by(rep, norm_ctrl_flag) %>%
  summarize(mean = mean(lfc_plasmid_vs_late),
            sd = sd(lfc_plasmid_vs_late)) %>%
  pivot_wider(names_from = norm_ctrl_flag, 
              values_from = c(mean, sd),
              names_glue = "{norm_ctrl_flag}_{.value}") %>%
  mutate(ssmd = (positive_control_mean - negative_control_mean) /
           (sqrt(positive_control_sd^2 + negative_control_sd^2)))

d.lfc_annot %>%
  filter(norm_ctrl_flag == "negative_control" | norm_ctrl_flag == "positive_control") %>%
  group_by(rep, norm_ctrl_flag) %>%
  summarize(mean = mean(lfc_plasmid_vs_late),
            var = var(lfc_plasmid_vs_late)) %>%
  pivot_wider(names_from = norm_ctrl_flag, 
              values_from = c(mean, var),
              names_glue = "{norm_ctrl_flag}_{.value}") %>%
  mutate(ssmd = (positive_control_mean - negative_control_mean) /
           (sqrt(positive_control_var + negative_control_var)))

```

### Replicate correlations

```{r}

results <- get_rep_cor_dfs(d.lfc_annot, "pgRNA_id", "lfc_plasmid_vs_late")

d.unadj_lfc_rep_cor_plot <- results[[1]]
d.unadj_lfc_rep_cor_summary <- results[[2]]

# d.unadj_lfc_rep_cor_plot

print_kbl(d.unadj_lfc_rep_cor_summary)

```

```{r}

## make replicate correlation plot
make_rep_cor_plot(d.unadj_lfc_rep_cor_plot, d.unadj_lfc_rep_cor_summary, "LFC")


```


```{r}

d.reps <- d.lfc_annot %>%
  ungroup() %>%
  dplyr::select(rep) %>%
  distinct()

d.lfc_rep_cor <- d.lfc_annot %>%
  ungroup() %>%
  dplyr::select(pgRNA_id, rep, lfc_plasmid_vs_late) %>%
  pivot_wider(names_from = rep,
              values_from = lfc_plasmid_vs_late) 
# d.lfc_rep_cor

d.cor <- d.lfc_rep_cor %>%
  dplyr::select(-pgRNA_id) %>%
  corrr::correlate() %>%
  shave() %>%
  stretch() %>%
  filter(!is.na(r)) %>%
  unite(c(x, y), col = "comparison", sep = "_vs_", remove = FALSE) %>%
  rename("sample1" = x, "sample2" = y)


comparisons <- d.cor %>% pull(comparison)

results <- lapply(comparisons, function(i){
  print(i)
  
  sample1 <- d.cor %>% filter(comparison == i) %>% pull(sample1)
  print(sample1)
  
  sample2 <- d.cor %>% filter(comparison == i) %>% pull(sample2)
  print(sample2)
  
  d.comparison_lfc <- d.lfc_rep_cor %>%
    dplyr::select(sample1, sample2) %>%
    rename("sample1" = sample1, "sample2" = sample2) %>%
    mutate("comparison" = i)
})

d.lfc_rep_cor_plot <- bind_rows(results)

d.lfc_rep_cor_plot %>%
  ggplot(aes(x = sample1, y = sample2)) + 
  geom_point(size = 1) +
  geom_density2d(aes(color = ..level..), alpha = 0.7) +
  geom_text(data = d.cor, 
            mapping = aes(x = -Inf, y = Inf, label = paste0("r=", round(r, 3))), 
            hjust = -0.25, vjust = 1.75, size = 3.5) +
  scale_colour_gradientn(colors = rev(contour_palette)) +
  labs(x = "sample1_LFC",
       y = "sample2_LFC", 
       color = "density") +
  plot_options +
  plot_theme +
  theme(aspect.ratio = square_ar) +
  facet_wrap(~comparison)

```

## Adjust LFCs

### Pos/neg controls

Take LFC, then subtract median of negative controls. This will result in the median of the nontargeting being set to 0. Then, divide by the median of negative controls (double non-targeting) minus median of positive controls (targeting 1 essential gene). This will effectively set the median of the positive controls (essential genes) to -1.

```{r}

d.control_group_medians <- d.lfc_annot %>%
  group_by(rep, norm_ctrl_flag) %>%
  filter(norm_ctrl_flag == "negative_control" | norm_ctrl_flag == "positive_control") %>%
  summarize(median_lfc = median(lfc_plasmid_vs_late))
print_kbl(d.control_group_medians)


```


```{r}

d.lfc_annot_adj <- d.lfc_annot %>%
  group_by(rep) %>%
  mutate(lfc_adj1 = lfc_plasmid_vs_late - median(lfc_plasmid_vs_late[norm_ctrl_flag == "negative_control"]),
         lfc_adj2 = lfc_adj1 / (median(lfc_adj1[norm_ctrl_flag == "negative_control"]) -
                                  median(lfc_adj1[norm_ctrl_flag == "positive_control"]))) %>%
  ungroup()
d.lfc_annot_adj

```


```{r}

## plot adjusted violin plot
# d.lfc_annot_adj %>%
#   ggplot(aes(x = norm_ctrl_flag, y = lfc_adj2, fill = norm_ctrl_flag)) +
#   geom_hline(yintercept = 0) +
#   geom_violin() +
#   geom_boxplot(outlier.shape = NA, coef = 0, width = 0.1) +
#   labs(x = "pgRNA_category",
#        y = "adjusted_lfc") +
#   plot_options +
#   plot_theme +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1),
#         legend.position = "none",
#         aspect.ratio = wide_ar) +
#   facet_wrap(~rep)

adjusted_norm_ctrl_violin_plot <- make_norm_ctrl_violin_plot(d.lfc_annot_adj, "lfc_adj2", "adjusted_LFC")

adjusted_norm_ctrl_violin_plot +
    facet_wrap(~rep)

```


### Expression

Since the pgPEN library uses non-targeting controls, we adjusted for the fact that single-targeting pgRNAs generate only two double-strand breaks (1 per allele), whereas the double-targeting pgRNAs generate four DSBs. To do this, we set the median (adjusted) LFC for unexpressed genes of each group to zero. 


#### Single-targeting
```{r}

d.lfc_annot_adj_single <- d.lfc_annot_adj %>%
  filter(target_type == "gene_ctrl" | target_type == "ctrl_gene") %>%
  ## make a flag variable to indicate which pgRNAs are targeting unexpressed
  ## single targets
  mutate(unexpressed_ctrl_flag = case_when(
    target_type == "gene_ctrl" & gene1_expressed_flag == FALSE ~ TRUE,
    target_type == "ctrl_gene" & gene2_expressed_flag == FALSE ~ TRUE,
    TRUE ~ FALSE 
  )) %>%
  group_by(rep) %>%
  mutate(lfc_adj3 = lfc_adj2 - median(lfc_adj2[unexpressed_ctrl_flag == TRUE]))

d.lfc_annot_adj_single_summary <- d.lfc_annot_adj_single %>%
  group_by(rep, unexpressed_ctrl_flag) %>%
  summarize(median = median(lfc_adj3))
print_kbl(d.lfc_annot_adj_single_summary)


```


```{r}

## plot before & after?

d.lfc_annot_adj_single %>%
  filter(!is.na(gene1_log2_tpm) | !is.na(gene2_log2_tpm)) %>%
  ggplot(aes(x = lfc_adj3, fill = unexpressed_ctrl_flag)) +
  geom_density(alpha = 0.7) +
  geom_vline(data = d.lfc_annot_adj_single_summary, 
             aes(xintercept = median, color = unexpressed_ctrl_flag),
             linetype = "dashed") +
  scale_fill_discrete(name = "expression_group", 
                      limits = c(FALSE, TRUE),
                      labels = c("expressed", "unexpressed")) +
  scale_color_discrete(name = "expression_group", 
                       limits = c(FALSE, TRUE),
                       labels = c("expressed", "unexpressed")) +
  labs(x = "adjusted_lfc") +
  plot_options +
  plot_theme +
  theme(aspect.ratio = wide_ar) +
  facet_wrap(~rep)

```

#### Double-targeting
```{r}

d.lfc_annot_adj_double <- d.lfc_annot_adj %>%
  filter(target_type == "gene_gene") %>%
  ## make a flag variable to indicate which pgRNAs are targeting double
  ## unexpressed targets
  mutate(unexpressed_ctrl_flag = case_when(
    gene1_expressed_flag == FALSE & gene2_expressed_flag == FALSE ~ TRUE,
    TRUE ~ FALSE)) %>%
  group_by(rep) %>%
  mutate(lfc_adj3 = lfc_adj2 - median(lfc_adj2[unexpressed_ctrl_flag == TRUE]))

d.lfc_annot_adj_double_summary <- d.lfc_annot_adj_double %>%
  group_by(rep, unexpressed_ctrl_flag) %>%
  summarize(median = median(lfc_adj3))
print_kbl(d.lfc_annot_adj_double_summary)

```

```{r}

d.lfc_annot_adj_double_plot <-  d.lfc_annot_adj_double %>% 
  filter(!is.na(gene1_log2_tpm) & !is.na(gene2_log2_tpm)) %>%
  mutate(n_genes_expressed = case_when(
      gene1_expressed_flag == FALSE & gene2_expressed_flag == FALSE ~ "0",
      gene1_expressed_flag == TRUE & gene2_expressed_flag == FALSE ~ "1",
      gene1_expressed_flag == FALSE & gene2_expressed_flag == TRUE ~ "1",
      gene1_expressed_flag == TRUE & gene2_expressed_flag == TRUE ~ "2")) 

d.lfc_annot_adj_double_plot_summary <- d.lfc_annot_adj_double_plot %>%
  group_by(rep, n_genes_expressed) %>%
  summarize(median = median(lfc_adj3))


```


```{r}

ggplot(d.lfc_annot_adj_double_plot, aes(x = lfc_adj3, fill = n_genes_expressed)) +
  geom_density(alpha = 0.7) +
  geom_vline(data = d.lfc_annot_adj_double_plot_summary,
             aes(xintercept = median, color = n_genes_expressed),
             linetype = "dashed") +
  labs(x = "adjusted_lfc") +
  plot_options +
  plot_theme +
  theme(aspect.ratio = wide_ar) +
  facet_wrap(~rep)

```

#### Re-join tbls
```{r}

### ntc_ntc
d.lfc_annot_adj_control <- d.lfc_annot_adj %>%
  filter(target_type == "ctrl_ctrl") %>%
  mutate(lfc_adj3 = lfc_adj2)

### single targeting
# colnames(d.lfc_annot_adj_single)

d.lfc_annot_adj_single <- d.lfc_annot_adj_single %>%
  dplyr::select(-unexpressed_ctrl_flag)

### double targeting
d.lfc_annot_adj_double <- d.lfc_annot_adj_double %>%
  dplyr::select(-unexpressed_ctrl_flag)

## bind rows
d.lfc_annot_adj_pgRNA <- bind_rows(d.lfc_annot_adj_double, d.lfc_annot_adj_single, d.lfc_annot_adj_control)

```

```{r}

## keep only columns of interest
d.lfc_annot_adj_pgRNA <- d.lfc_annot_adj_pgRNA %>%
  dplyr::select(pgRNA_id, rep, paralog_pair, lfc_adj3, target_type:gene2_essential_flag) %>%
  ## rename final adjusted column to CRISPR_score
  rename(CRISPR_score = lfc_adj3)

```


### Save pgRNA output
```{r}

save_tbl(d.lfc_annot_adj_pgRNA)

```

## Target-level values

```{r}

d.lfc_annot_adj_target <- d.lfc_annot_adj_pgRNA %>%
  group_by(rep, pgRNA_target) %>%
  mutate(target_mean_CS = mean(CRISPR_score),
         target_median_CS = median(CRISPR_score)) %>%
  distinct(pgRNA_target, .keep_all = TRUE) %>%
  dplyr::select(-c(pgRNA_id, CRISPR_score, contains("seq")))
  

```

```{r}

results <- get_rep_cor_dfs(d.lfc_annot_adj_target, "pgRNA_target", "target_mean_CS")

d.adj_mean_lfc_rep_cor_plot <- results[[1]]
d.adj_mean_lfc_rep_cor_summary <- results[[2]]

# d.unadj_lfc_rep_cor_plot

print_kbl(d.adj_mean_lfc_rep_cor_summary)

```

```{r}

## make replicate correlation plot
make_rep_cor_plot(d.adj_mean_lfc_rep_cor_plot, d.adj_mean_lfc_rep_cor_summary, "target_mean_CS")


```


```{r}

## target-level violin plot
adjusted_norm_ctrl_violin_plot <- make_norm_ctrl_violin_plot(d.lfc_annot_adj_target, "target_mean_CS", "target_mean_adjusted_LFC")

adjusted_norm_ctrl_violin_plot +
    facet_wrap(~rep)

```

```{r}

## mean across reps violin plot
d.lfc_annot_adj_target_rep_mean <- d.lfc_annot_adj_target %>%
  group_by(pgRNA_target) %>%
  mutate(rep_target_mean_CS = mean(target_mean_CS)) %>%
  distinct(pgRNA_target, .keep_all = TRUE) %>%
  dplyr::select(-c(rep, target_mean_CS, target_median_CS))


rep_mean_adjusted_norm_ctrl_violin_plot <- make_norm_ctrl_violin_plot(d.lfc_annot_adj_target_rep_mean, "rep_target_mean_CS", "mean_adjusted_LFC_across_reps")

rep_mean_adjusted_norm_ctrl_violin_plot

# adjusted_norm_ctrl_violin_plot

```


```{r}

## get (& plot) mean across retained reps
d.lfc_annot_adj_pgRNA_rep_mean <- d.lfc_annot_adj_pgRNA %>%
  group_by(pgRNA_id) %>%
  mutate(rep_mean_CS = mean(CRISPR_score)) %>%
  distinct(pgRNA_target, .keep_all = TRUE) %>%
  dplyr::select(-c(rep, CRISPR_score))


rep_mean_adj_pgRNA_norm_ctrl_violin_plot <- make_norm_ctrl_violin_plot(d.lfc_annot_adj_pgRNA_rep_mean, "rep_mean_CS", "mean_adjusted_LFC_across_reps")

rep_mean_adj_pgRNA_norm_ctrl_violin_plot

```



