---
title: "GI mapping pipeline: pgRNA counts QC"
author: "Phoebe Parrish"
date: "`r format(Sys.time(), '%m/%d/%y %I:%M %p')`"
output: html_document
params: 
  cell_line: "HeLa"
  base_dir: "/Volumes/Projects/paralog_pgRNA/pgPEN_library/GI_mapping"
---


## To Do
* convert sample list to factor - make sure there is a consistent naming scheme in the barcode ref file, then use that somehow? Sort it alphabetically?
* write a function to summarize n and percent of TRUE values, return n(TRUE)
* allow user to input cutoff used to filter reads


## Setup
```{r, include = FALSE, eval = FALSE}

## possible params for future use: 
# params: 
#   counts_file: 
#     value: x
#   output_dir: 
#     value: x
# knit: (function(inputFile, encoding) {
#     rmarkdown::render(inputFile, 
#       encoding = encoding, 
#       output_dir = "`r params$output_dir`")
#   })
# knit: function(inputFile, encoding){
#   rmarkdown::render(inputFile, 
#     encoding = encoding, 
#     output_dir = "../results/pgRNA_counts_QC")}

```


```{r, include = FALSE, echo = FALSE, cache = FALSE}

knitr::opts_chunk$set(
  results = "hold"
)

# knitr::knit_hooks$set(output = function(x, options){
#   gsub("```\n*```r*\n*", "", x)
# })

```

### Load packages
```{r setup, results = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(tidylog)
library(pheatmap) # for making correlation heatmap
library(RColorBrewer) # for heatmap colors
# library(raster) # for calculating %CV
library(kableExtra) # for formatting kables

```

### Save variables and functions
```{r}

## get functions and vars from shared R script
source(file.path(params$base_dir, "workflow", "scripts", "shared_functions_and_variables.R"))

contour_palette <- colorRampPalette(brewer.pal(n = 9, name ="Spectral"))(50)

```


```{r}

## convert to Snakemake input later
in_dir <- file.path(params$base_dir, "results", "pgRNA_counts")

out_dir <- file.path(params$base_dir, "results", "pgRNA_counts_QC")

```

### Read in files

```{r, results = FALSE, message = FALSE, warning = FALSE}

d.counts <- read_tsv(file.path(in_dir, paste0("PP_pgPEN_", params$cell_line, "_counts.txt")), col_names=TRUE)

```


## QC

### Reformat d.counts
```{r}

## what does d.counts look like?
print_kbl(head(d.counts, 10))

```

```{r}

## pivot_longer to tidy format
d.counts_long <- d.counts %>%
  dplyr::select(-c(seq_1, seq_2)) %>%
  pivot_longer(!id, 
               names_to = "sample", 
               values_to = "count")

## how many reads are in each sample?
d.summ <- d.counts_long %>%
  group_by(sample) %>%
  summarize(sum = sum(count), 
            coverage = round((sum / 31700), 2))
print_kbl(d.summ)

## convert sample to a factor 
## NOTE: figure out a way to do this so it automatically sorts logically
d.counts_long <- d.counts_long %>%
  mutate(sample = factor(sample)) 

```


### CDF 
```{r}

d.counts_cdf <- d.counts_long %>%
  group_by(sample) %>%
  mutate(count_norm = -log10((count+1)/sum(count)))

pgRNA_counts_cdf <- ggplot(d.counts_cdf, aes(x = count_norm, color = sample)) +
  stat_ecdf() +
  labs(x = "-log10(count/total_count)", # bquote(~-log[10]~"(count/total_count)")
       y = "Expected_pgRNAs",
       color = "Sample") +  
  plot_options +
  plot_theme +
  theme(aspect.ratio = wide_ar)
pgRNA_counts_cdf
# save_plot(pgRNA_counts_cdf)

```

### CPM for each sample
```{r}

## plot counts per million for each sample
d.counts_cpm <- d.counts_long %>%
  group_by(sample) %>%
  mutate(cpm = (((count)/sum(count))*1e6)) %>%
  mutate(log2_cpm = log2(cpm +1))

n_samples <- d.counts_long %>%
  distinct(sample) %>%
  nrow()

sample_cpm_histogram <- ggplot(d.counts_cpm, aes(x = log2_cpm, fill = sample)) +
  geom_histogram(color = "black", binwidth = 0.5) +
  plot_options +
  plot_theme +
  theme(aspect.ratio = wide_ar,
        legend.position = "none") +
  facet_wrap(~sample, scales = "free_y", ncol = ceiling(n_samples/2))
sample_cpm_histogram
# save_plot(sample_cpm_histogram)

```

### Sample corr heatmap

```{r}

## sample correlation heatmap
d.counts_cpm_cor <- d.counts_cpm %>%
  dplyr::select(id, sample, cpm) %>%
  pivot_wider(names_from = "sample",
              values_from = "cpm") %>%
  dplyr::select(-id) %>%
  cor() %>%
  round(2) %>%
  data.frame()
print_kbl(d.counts_cpm_cor)

```


```{r}

colors <- colorRampPalette(brewer.pal(n = 9, name ="YlGnBu"))(50)

sample_cor_heatmap_unfiltered <- pheatmap(d.counts_cpm_cor,
                 col = colors,
                 border_color = "white",
                 cellwidth = 20, cellheight = 20,
                 treeheight_row = 20, treeheight_col = 20,
                 ## extra stuff
                 cluster_rows = TRUE,
                 cluster_cols = TRUE,
                 cex = 1, clustering_distince_rows = "euclidean",
                 cex = 1, clustering_distance_cols = "euclidean",
                 cluster_method = "complete")
sample_cor_heatmap_unfiltered

# png(file.path(out_dir, "plots", "png", "sample_correlation_heatmap_unfiltered.png"), 
#     units = "in", res = 300, 
#     width = 4.51, height = 4.51)
# pdf(file.path(out_dir, "plots", "pdf", "sample_correlation_heatmap_unfiltered.pdf"), 
#     width = 4.51, height = 4.51)
sample_cor_heatmap_unfiltered
# dev.off()

```



### Flag low counts

```{r}

## flag pgRNAs with count = 0 at any time point
d.counts_cpm_filter <- d.counts_cpm %>%
  group_by(id) %>%
  mutate(zero_count = case_when(
    any(count == 0) ~ TRUE, ## if any value in the group = 0, set value to TRUE
    TRUE ~ FALSE)) %>% ## if above condition is not met, set value to FALSE
  ungroup()

## how many guides will be removed using this filter? (zero_count == TRUE)
d.summ <- d.counts_cpm_filter %>%
  dplyr::select(id, zero_count) %>%
  distinct(id, .keep_all = TRUE) %>%
  group_by(zero_count) %>%
  summarize(n = n()) %>%
  mutate(percent = round(((n/sum(n))*100), 2))
print_kbl(d.summ)

```


```{r}

## flag pgRNAs with low plasmid read counts

## filter pgRNA df for plasmid reads only
d.counts_cpm_plasmid <- d.counts_cpm %>%
  filter(sample == "Day00_RepA")  ## change to Day00 once I figure out a good naming scheme

## what does the plasmid read count distribution look like? 
plasmid_cpm_histogram <-  ggplot(d.counts_cpm_plasmid, aes(x = log2_cpm)) +
  geom_histogram(binwidth = 0.2,
                 color = "black", 
                 fill = "gray60") +
  plot_options +
  plot_theme +
  theme(aspect.ratio = wide_ar)
plasmid_cpm_histogram
# save_plot(plasmid_cpm_histogram)

## determine plasmid cutoff
d.counts_cpm_plasmid_stats <- d.counts_cpm_plasmid %>%
  summarize(median = median(log2_cpm),
            Q1 = quantile(log2_cpm, probs = 0.25),
            Q3 = quantile(log2_cpm, probs = 0.75),
            lower_outlier = (Q1 - 1.5*(Q3 - Q1)))
print_kbl(d.counts_cpm_plasmid_stats)

## save your selected cutoff as a variable
plasmid_cpm_cutoff <- d.counts_cpm_plasmid_stats %>%
  pull(lower_outlier) %>%
  unlist() %>%
  unname()

```


```{r}

## add your cutoff line to the plasmid plot
plasmid_cpm_histogram_cutoff <- plasmid_cpm_histogram + 
  geom_vline(xintercept = plasmid_cpm_cutoff, ## adjust based on selected cutoff
             linetype = "dashed") 
plasmid_cpm_histogram_cutoff
# save_plot(plasmid_cpm_histogram_cutoff)

## add filter variable to pgRNA df based on selected cutoff
d.counts_cpm_plasmid_filter <- d.counts_cpm_plasmid %>%
  mutate(low_plasmid_cpm = case_when(
    log2_cpm < plasmid_cpm_cutoff ~ TRUE, ## if plasmid log2_cpm < cutoff, set to TRUE
    TRUE ~ FALSE)) %>% ## if above condition is not met, set to FALSE
  ungroup()

## how many guides will be removed based on this filter?
d.summ <- d.counts_cpm_plasmid_filter %>%
  dplyr::select(id, low_plasmid_cpm) %>%
  distinct(id, .keep_all = TRUE) %>%
  group_by(low_plasmid_cpm) %>%
  summarize(n = n()) %>%
  mutate(percent = round(((n/sum(n))*100), 2))
print_kbl(d.summ)

```

```{r}

## add low_plasmid_cpm variable to filter DF
d.counts_cpm_filter <- d.counts_cpm_plasmid_filter %>%
  dplyr::select(id, low_plasmid_cpm) %>%
  right_join(d.counts_cpm_filter, by = "id") %>%
  dplyr::select(id, sample:zero_count, low_plasmid_cpm) %>% ## reorder cols
  ungroup()

```


```{r}

## how many pgRNAs will be removed by both filters?
d.counts_cpm_filter <- d.counts_cpm_filter %>%
  dplyr::select(id, zero_count, low_plasmid_cpm) %>%
  distinct(id, .keep_all = TRUE) %>%
  group_by(id) %>%
  mutate(rm_pgRNA = case_when(
    any(zero_count == TRUE | low_plasmid_cpm == TRUE) ~ TRUE,
    TRUE ~ FALSE)) %>%
  ungroup()

## write a function to do this summary? and return n(TRUE)?
d.summ <- d.counts_cpm_filter %>%
  group_by(rm_pgRNA) %>%
  summarize(n = n()) %>%
  mutate(percent = round(((n/sum(n))*100), 2))
print_kbl(d.summ) 

```


```{r}

## confirm that n removed pgRNAs = # pgRNAs with low plasmid count + # pgRNAs with 0 reads at any time point
d.summ_complete <- d.counts_cpm_filter %>%
  mutate(flag_group = case_when(
    zero_count == TRUE & low_plasmid_cpm == TRUE ~ "both",
    zero_count == TRUE & low_plasmid_cpm == FALSE ~ "zero_count_only",
    low_plasmid_cpm == TRUE & zero_count == FALSE ~ "low_plasmid_cpm_only",
    low_plasmid_cpm == FALSE & zero_count == FALSE ~ "neither",
    TRUE ~ "error"
  )) %>%
  group_by(flag_group) %>%
  summarize(n = n())
print_kbl(d.summ_complete) 

```


## Write output
```{r}

## add counts back into flagged df, write output
d.counts_cpm_flag_long <- left_join(d.counts_cpm, d.counts_cpm_filter, by = "id")
save_tbl(d.counts_cpm_flag_long)

d.counts_cpm_flag_wide <- d.counts_cpm_flag_long %>%
  pivot_wider(names_from = sample, 
              values_from = count:log2_cpm,
              names_glue = "{sample}_{.value}")
save_tbl(d.counts_cpm_flag_wide)

```


```{r, include = FALSE}

knitr::knit_exit()

```

